Site-Heterogeneous Character Change Models for Morphology

### 2019-01-09 Papers/Draft_SHDM.tex
add changes to text

\documentclass[]{article}
\usepackage{amsmath} 
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{caption}
\begin{document}

\title{Site-Heterogeneous Character Change Models for Morphology}
\author{Wright, Pett, Students, Heath}
\date{Today}
\maketitle

@@

Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
The Mk model is the most commonly-used model for incorporating morphology in maximum likelihood and Bayesian work.
This model is a generalization of the Jukes-Cantor model, assumes that there is an equal rate of forward and backwards transitions in morphological traits.
In this study, we explore the use of models that allow asymmetrical transition rates.
We use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
In particular, we examine the use of two models to allow the equilibrium character frequency to vary. 
One model, a Beta prior on state frequencies is similar to a model implemented in MrBayes and can be used to allow asymmetrical transition rates in binary characters.
We also describe a Dirichlet model to allow asymmetrical transitions in multistate characters.
These two models can be used to explicitly model different character frequencies within one dataset, as opposed to assuming a single common mechanism with respect to among character frequency variation.

@@

	Recent interest in using Bayesian methods to model morphological evolution to estimate phylogenetic trees has spurred many studies into how well the existing phylogenetic models perform, particularly in relation to parsimony methods.
	However, there has been little work expanding the toolkit of methods available to researchers estimating trees from morphological data. 
	The most common way to perform likelihood or Bayesian phylogenetic analysis using discrete morphological data is by applying a model of morphological evolution called the Mk model.

@@

	As a generalization of the Jukes-Cantor model, it makes the same set of assumptions: that change between any two or more states  is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.

@@

	If a matrix was coded with respect to specific outgroups, often the outgroup will mostly have a `0' state, with ingroup taxa having typically having state `1' or higher, implying a highly one-directional transition matrix.

@@

	This lack of common meaning limits a researcher's ability to specify a model to apply to the whole matrix.
	This complicates researchers' ability to specify unique $\alpha$ parameters to the Q-matrix, or to apply multiple Q-matrices across the dataset.
	While the basic model is very simple, some elaborations of the model are common.

@@

@@

In most likelihood models of character or sequence evolution, the probability of observing a change in a character is dependent not only on the probability of change from this character to another, but on the frequency of the starting character.
Even if a character has a low probability of changing, change in that character may still be observed many times if the stationary frequency of that character is high (i.e., the character state is observed many times). 
Likewise, a highly probable changes may be seen relatively rarely if the starting character is observed rarely. 

@@

In RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of MrBayes, and a constrained version of the CAT model. 

@@

A mathematical descriptionf of this model can be found in the methods section. \par

@@

This dataset has interesting properties.

@@

We would expect for these characters to be better modeled by a model that can accommodate asymmetrical transition rates.
Some apomorphies of the ant group are also gained after the divergence of the stem lineages, which also violates the assumption of equal change probabilities.

@@

\subsubsection{Discretized beta model}
We model site-specific binary state frequencies by assuming the state frequencies $\pi_i$ are drawn from a 2-dimensional Dirichlet (Beta) prior distribution with hyperparameters $\alpha,\beta$.
Then, the likelihood of the data $D$ is computed by marginalizing over the value for $\pi_i$ at each site.

@@

\subsection{Data Matrices}
\subsubsection{Empirical Matrices}
Several large and well-documented ant matrices were used in this study.
The first was that of Keller (2011). 
This matrix is of extant ant groups. 
Keller's matrix was collected with special attention to the poneromorph subgroups (Amblyoponinae, Ectatomminae,
Heteroponerinae, Paraponerinae, Ponerinae, and Proceratiinae) and has 139 characters and 105 taxa.
The matrix is highly complete for a morphological dataset, with about 60\% of cells being filled.
Of the total character set, 100 characters were binary and 39 characters were multistate.
Because of the scope of Keller's study, the taxon sampling is biased towards the poneromorphs, and away from the other ant subfamilies, including large subfamilies such as the Dolichoderinae and the Formicinae. \par
We also used data from Barden and Grimaldi (2015). 
This matrix contained both extant and extinct ants, and was collected in order to place stem ants from the Cretaceous period.
One crown ant amber fossil, \textit{Kyromyrma neffi}, was included in this matrix, but the key feature of this matrix is the sampling of the stem group, including multiple samples from the \textit{Gerontoformica} genus.
This matrix also expands sampling in the non-ponerine groups that are underrepresented in the Keller matrix.
The Barden-Grimaldi matrix contained 42 characters and 41 taxa. 
Of these characters, 26 were binary.\par
In order to achieve maximum coverage, we merged these character matrices.
The taxonomic overlap between the two was 11 taxa, and all subfamilies of the Formicidae except the extinct Formiciinae were represented in the matrix. 
Eleven characters were represented in both matrices without any recoding, and both matrices had many characters that were inapplicable to the other matrices.
Inapplicable characters were left in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
Six characters were recoded to make the character states uniform between the two matrices.
Most of these changes simply involved changing the terminology used in reference to the character states.
For example, Barden and Grimaldi's character 15 is the same as Keller's character 47, but the two matrices had inverse character codings relative to one another.
The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. 
We will refer to this dataset as the combined dataset. \par
We modeled all three datasets according to both models.
First, to test the Beta model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
This dataset was chosen because the dataset size and completeness is very typical to a morphological dataset, particularly one involving fossils.
Because of the presence of stem ants, we also expect to see strong violations of the assumptions of the Mk model in this dataset.
We also used the combined matrix, stripped of multistate characters to test the model to look at the effect of estimating a larger tree using more characters.
In both cases, we performed Bayes Factor model fitting to determine how many discrete Beta categories best model the data. \par
To test the full SHDM model, we used both the Barden-Grimaldi dataset and the combined dataset.
In testing the SHDM model, we modeled the binary and multistate characters together.
This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data (both more taxa and more characters) to the phylogenetic question. 
\par
All estimations were performed in the software RevBayes.
\subsubsection{Model comparison}

@@

The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 2C and 2D.
The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 2A, and including the multistate characters can be seen in 2B.
In the Barden datasets, the data support 4 character state frequency variation categories for both the binary and multistate data.
In the combined dataset, 6 categories are supported for binary data and 7 are supported for multistate. \par
When a phylogeny is estimated from the Barden dataset under the supported model, it is largely unresolved.

@@

Why then would a model statistically supported to be a better fit to the data not lead to a more resolved tree?

@@

However, in such a large and diverse group, not all homoplasy can be avoided, and previous phylogenetic and taxonomic work in the group has also reported a lack of support for well-supported molecular clades due to a lack of morphological synapomorphies among these groups. 

@@

@@

If, for small matrices, a model that is better fit does not yield a more resolved tree, what does this mean for morphologists? 

@@

Pragmatically, prior work has supported that model choice is more important when the problem is hard \citep{wright2014, Wright2016}. 

@@

\bibliographystyle{sysbio}
\bibliography{bibliography.bib}

\begin{figure}
  \caption{A. A graphical model demonstrating the discrete Beta model. B. A graphical model displaying the SHDM. Model realization following Hoehna et al. 2016.}
    \includegraphics[scale=0.35]{fig/1AB}
\end{figure}

\begin{figure}
  \caption{Marginal likelihood calculations for the eight models tested for teach of the four empirical datasets. The number of categories into which the Beta (binary data) or Dirichlet (multistate data) is discretized is indicated on the x-axis. Note that due to differing dataset sizes, each plot has a unique y-axis.}
    \includegraphics[scale=0.5]{fig/SS_plot_nostar}
\end{figure}

@@

\begin{figure}
  \caption{Accuracy of phylogenetic estimation in small large (135 taxa, 163 character) simulated multistate datasets under the SHDM with four transition rate asymmetry categories. Both over- and underparameterization appear to be detrimental to accuracy, though underparamterization is worse.} 
    \includegraphics{fig/LargeMultiState}
\end{figure}

>>>>

\documentclass[]{sysbio}
\usepackage{amsmath} 
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{caption}
\begin{document}
\title{Site-Heterogeneous Character Change Models for Morphology}
\author{Wright, Pett, Students, Heath}
\date{Today}
\maketitle

@@

Morphological data is challenging to model due to the complexity of morphological characters relative to nucleotide sequence or amino acid characters. 
Particularly problematic is that morphological characters are coded in a way that is meaningful to human observers, but that this does not necessarily equate to a shared evolutionary mechanism among characters.
Character states may vary in their meaning between characters, and  the magnitude of underlying genetic changes required to produce a morphological change will also vary among characters.
The Mk model is the most commonly-used model for incorporating morphology in maximum likelihood and Bayesian phylogenetic estimation.
This model is a generalization of the Jukes-Cantor model which assumes that there is an equal rate of forward and backwards transitions in morphological traits.
In this study, we explore the use of models that allow asymmetrical transition rates.
We use an expanded combined set of two ant (Formicidae) data matrices, composed of one matrix of extinct and extant ants and one matrix of extant ants to estimate a phylogeny for the group.
There is extensive sampling in the stem ants, which leads to classes of gains and loss within the data, as ant apomorphies are gained, but characters more similar to the outgroup are lost. 
This violates the assumption of equal transition rates among sites.
In particular, we examine the use of two models to allow the equilibrium character frequency to vary. 
One model, a Beta prior on state frequencies is similar to a model implemented in MrBayes and can be used to allow asymmetrical transition rates in binary characters.
We also describe a Dirichlet model to allow asymmetrical transitions in multistate characters.
These two models can be used to explicitly model different character frequencies within one dataset, as opposed to assuming a single common mechanism with respect to among character frequency variation.

@@

	Recent interest in using Bayesian methods to model morphological evolution for the estimation of phylogenetic trees has spurred many studies into how well the existing phylogenetic models perform, particularly in relation to parsimony methods.
	However, there has been little work expanding the toolkit of methods available to researchers estimating trees from morphological data \citep{parins2017, dosreis2018, Alvarez-Carretero441105}. 
	The most common way to perform likelihood or Bayesian phylogenetic analysis using discrete morphological data is by applying a model of morphological evolution called the Mk model \citep{lewis2001}.

@@

	As a generalization of the Jukes-Cantor model \citep{Jukes1969}, it makes the same set of assumptions: that change between any two or more states  is equally likely, that the stationary character frequencies of every character are the same, and that each character is always in one of \textit{k} known states.

@@

	If a matrix was coded with respect to specific outgroups, characters in the outgroup taxa will mostly have `0' character states, with ingroup taxa having typically having state `1' or higher, implying a highly one-directional transition matrix.

@@

	This lack of common meaning may strike a researcher as evidence that a single common mechanism model, such as the Mk model, cannot be applied to the dataset.
	It also complicates researchers' ability to specify unique $\alpha$ parameters to the Q-matrix, or to apply multiple Q-matrices across the dataset, if there is little information by which to model characters separately.

@@

	While the basic Mk model is very simple, some elaborations of the model are common.
	Models such as a likelihood-based Dollo model have also been implemented. 
	Morphological models also accommodate among-site rate variation (ASRV), as in molecular data.
	This is typically achieved via the Gamma distribution, though there is evidence than lognormal distributions may better reflect morphological data, as invariant and very high-rate characters are often excluded from matrices \citep{Harrison2014} \par

@@

In most likelihood models of character or sequence evolution, the probability of observing a change from one character state is dependent not only on the probability of change from this character state to another, but on the frequency of the starting character state.
Even if a character state  has a low probability of changing to another state (low exchangeability), this change may still be observed many times if the stationary frequency of that character state is high (i.e., the starting character state is observed many times). 
Likewise, a highly probable changes may be seen relatively rarely if the starting character state is observed rarely. 

@@

In the software RevBayes, we have implemented a number of useful extensions to the Dirichlet prior of MrBayes, and a constrained version of the CAT model. 

@@

A mathematical description of this model can be found in the methods section. \par

@@

Barden and Grimaldi's dataset has interesting properties.

@@

We would expect for these characters to be better described by a model that can accommodate asymmetrical transition rates.
Some apomorphies of the ant group are also gained after the divergence of the stem lineages and not lost frequently, which also violates the assumption of equal change probabilities.

@@

\subsubsection{Discretized Beta model}
We model site-specific binary state frequencies by assuming the state frequencies $\pi_i$ are drawn from a 2-dimensional Dirichlet (Beta) prior distribution with hyperparameters $\alpha,\beta$.
Then, the likelihood of the data $D$ is computed by marginalizing over the value for $\pi_i$ at each site.

@@

\subsection{Data Matrices}
\subsubsection{Empirical Matrices}
Several large and well-documented ant matrices were used in this study.
The first was that of Keller (2011). 
This matrix is of extant ant groups. 
Keller's matrix was collected with special attention to the poneromorph subgroups (Amblyoponinae, Ectatomminae,
Heteroponerinae, Paraponerinae, Ponerinae, and Proceratiinae) and has 139 characters and 105 taxa.
The matrix is highly complete for a morphological dataset, with about 60\% of cells being filled.
Of the total character set, 100 characters were binary and 39 characters were multistate.
Because of the scope of Keller's study, the taxon sampling is biased towards the poneromorphs, and away from the other ant subfamilies, including large subfamilies such as the Dolichoderinae and the Formicinae. \par
We also used data from Barden and Grimaldi (2015). 
This matrix contained both extant and extinct ants, and was collected in order to place stem ants from the Cretaceous period.
One crown ant amber fossil, \textit{Kyromyrma neffi}, was included in this matrix, but the key feature of this matrix is the sampling of the stem group, including multiple samples from the \textit{Gerontoformica} genus.
This matrix also expands sampling in the non-ponerine groups that are underrepresented in the Keller matrix.
The Barden-Grimaldi matrix contained 42 characters and 41 taxa. 
Of these characters, 26 were binary.\par
In order to achieve maximum coverage, we merged these character matrices.
The taxonomic overlap between the two was 11 taxa, and all subfamilies of the Formicidae except the extinct Formiciinae were represented in the matrix. 
Eleven characters were represented in both matrices without any recoding, and both matrices had many characters that were inapplicable to the other matrices.
Characters that are inapplicable in parts of the tree were retained in the matrix due to their ability to resolve bipartitions in the the authors' respective groups of interest. \par
Six characters were recoded to make the character states uniform between the two matrices.
Most of these changes simply involved changing the terminology used in reference to the character states.
For example, Barden and Grimaldi's character 15 is the same as Keller's character 47, but the two matrices had inverse character codings relative to one another.
The remainder of the changes were changing binary characters to a multistate in cases where one author had more states than the other. 
We will refer to this dataset as the combined dataset. \par
We modeled all three datasets according to both models.
First, to test the Beta model on binary-only datasets, we used Barden and Grimaldi's dataset with the multistate characters removed.
This dataset was chosen because the dataset size and completeness is very typical to a morphological dataset, particularly one involving fossils.
Because of the presence of stem ants, we also expect to see strong violations of the assumptions of the Mk model in this dataset.
We also used the combined matrix, stripped of multistate characters to test the model to look at the effect of estimating a larger tree using more characters.
In both cases, we performed Bayes Factor model fitting to determine how many discrete Beta categories best model the data. \par
To test the full SHDM model, we used both the Barden-Grimaldi dataset and the combined dataset.
In testing the SHDM model, we modeled the binary and multistate characters together.
This allowed us to both examine the impact of allowing character frequencies to be drawn from a Dirichlet, and to examine the effect of adding more data (both more taxa and more characters) to the phylogenetic question. 
\par
All estimations were performed in the software RevBayes.
\subsubsection{Model comparison}

@@

The marginal likelihood comparisons for the Barden and Grimaldi dataset can be seen in Fig. 2C and 2D.
The marginal likelihood comparisons for the combined dataset with the multistate characters removed can be seen in Fig. 2A, and including the multistate characters can be seen in 2B.
In the Barden datasets, the data support 4 character state frequency variation categories for both the binary and multistate data.
In the combined dataset, 6 categories are supported for binary data and 7 are supported for multistate. \par
When a phylogeny is estimated from the Barden dataset under the supported model, it is largely unresolved. 
The tree (Fig. 3) supports the monophyly of crown ants. 
Haidomyrmicine ants are monophyletic, and less closely related to the crown ants than the Sphecomyrmine ants are, a relationship also supported by Barden and Grimaldi. 
Many relationships in the crown ants supported by Barden and Grimaldi, such as monophyletic Ponerine, Mymicine, and Formicine clades are also supported in our analysis.  
We also find support for an assemblage of Doryline, Leptilline and other ants found by the authors, though the exact relationships are different, and poorly-supported. 
Much as in Barden and Grimaldi's unconstrained analysis, many subfamilies are monophyletic, as expected from molecular analysis, but their relationships to one another remain vague and poorly supported. 
\par

Combined dataset results go here. \par

@@

Even with strongly supported models, the tree is still poorly resolved.

@@

However, in such a large and diverse group, not all homoplasy can be avoided, and previous phylogenetic and taxonomic work in the group has also reported a lack of support for well-supported molecular clades due to a dearth of morphological synapomorphies among these groups. 

@@

However, our methods do reasonably well at teasing apart homoplasy from phylogenetic signal.
For example, \textit{Martialis Heureka}, an ant that is the only known member of the subfamily Martialinae \citep{rabeling2008}, is supported as sister to \textit{Amblyopone pallipes}, despite elongate mandibles strongly reminiscent of stem ants, such as the Haidonmyrmecine ants.

@@

If, for small matrices, a model that is better fit does not yield a more resolved tree, or stronger support for the bifurcations that are discovered, what does this mean for morphologists? 

@@

Pragmatically, prior work has supported that model choice is more important when the problem is challenging \citep{wright2014, Wright2016}. 

@@

\bibliographystyle{sysbio.bst}
\bibliography{bibliography}

\begin{figure}
  \caption{A. A graphical model demonstrating the discrete Beta model. B. A graphical model displaying the SHDM. Model realization following Hoehna et al. 2016.}
    \includegraphics[width=\textwidth]{fig/1AB}
\end{figure}

\begin{figure}
  \caption{Marginal likelihood calculations for the eight models tested for teach of the four empirical datasets. The number of categories into which the Beta (binary data) or Dirichlet (multistate data) is discretized is indicated on the x-axis. Note that due to differing dataset sizes, each plot has a unique y-axis.}
    \includegraphics[width=\textwidth]{fig/SS_plot_nostar}
\end{figure}

\begin{figure}
  \caption{Strict consensus tree estimated from the dataset of Barden and Grimaldi (2015) under the SHDM. } 
    \includegraphics[scale=0.75]{fig/bartree}
\end{figure}

@@

\begin{figure}
  \caption{Accuracy of phylogenetic estimation in sma	ll large (135 taxa, 163 character) simulated multistate datasets under the SHDM with four transition rate asymmetry categories. Both over- and underparameterization appear to be detrimental to accuracy, though underparamterization is worse.} 
    \includegraphics{fig/LargeMultiState}
\end{figure}
